Characteristic polynomials of p-adic matrices

### 2017-04-21 charpoly/charpoly.tex
Section 2.1 rewritten

@@

\item \done{In section 3, Algorithm 1, we have clarified why line 5 does not take quartic time} \\
      \david{Tristan, can you check that my clarification before Proposition 3.2 is correct?} \\
      \tristan{This is perfect.}

@@

\item \tdo{In the abstract, clarify what kind of operations are being counted.}
\item \tdo{In the abstract, clarify what we mean by ``precision exactly $O(p^k)$.''}\\
      \xavier{By the way, we should also mention in the abstract the $\softO(n^\omega)$ algorithm.}

@@

\item \tdo{Please define lattice in the tangent space and provide a simple, illustrative example.}\\
      \xavier{Maybe we should instead remove any mention to the tangent space: it's not really necessary.}\\
      \david{That's probably a good idea.  We'll have to make a remark about ignoring the leading monic term.}

@@

\xavier{I propose to expand the first paragraph of Section 2.1 (and even
maybe make a separate subsection for it). Concretely, I suggest to 
(1)~do not talk about the tangent space (but instead consider lattices
in $M_n(K)$),
(2)~give precision definition of flat, jagged precisions and relate them 
with lattices,
(3)~maybe cite the notes of my recent lecture at JNCF in which all of
this is detailed.}

@@

Instead, we separate the computation of the precision of $\chi_M$ from the computation
of an approximation to $\chi_M$.  Given some precision on $M$, we use \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}
to find the best possible precision for $\chi_M$.  The analysis of this precision is the subject
of much of this paper.  With this precision known, the actual calculation of $\chi_M$
may proceed by lifting $M$ to a temporarily higher precision and then using a sufficiently
stable algorithm (see Remark \ref{rem:lift_for_optimal}).

@@

\todo{Rewrite this subsection with more details.}

We recall some of the definitions and results of \cite{caruso-roe-vaccon:14a}
as a foundation for our discussion of the precision for the characteristic polynomial
of a matrix.  We will be concerned with two $K$-manifolds in what follows:
the space $M_n(K)$ of $n \times n$ matrices with entries in $K$ and the space
$K_n[X]$ of monic degree $n$ polynomials over $K$.  Given a matrix $M \in M_n(K)$,
the most general kind of precision structure we may attach to $M$ is a
\emph{lattice} $H$ in the tangent space at $M$.  However, representing an
arbitrary lattice requires $n^2$ basis vectors, each with $n^2$ entries.  We therefore
frequently work with certain classes of lattices, either \emph{jagged} lattices
where we specify a precision for each matrix entry or \emph{flat} lattices where
every entry is known to a fixed precision $O(p^N)$.  Similarly, precision for
monic polynomials can be specified by giving a lattice in the tangent space
at $f(X) \in K_n[X]$, or restricted to jagged or flat precision in the interest
of simplicity.

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
For a matrix $M \in M_n(K)$, we identify the tangent space $V$
at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi_M$ with
the space $K_{<n}[X]$ of polynomials of degree less than $n$.
Let $\adj(M)$ denote the adjugate\footnote{In \citelist{\cite{caruso-roe-vaccon:14a} \cite{caruso-roe-vaccon:15a}} the adjugate was referred to using the non-standard term \emph{comatrix}} of $M$ (when $M \in \GL_n(K)$,
we have $\adj(M) = \det(M) M^{-1}$) and $d\chi_M$ the differential at $M$.  Recall
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
that $d\chi_M$ is given by
\begin{equation} \label{eq:dchi}
d\chi_M: dM \mapsto \tr(\adj(X-M) \cdot dM).
\end{equation}

\xavier{I propose to move Proposition \ref{prop:mainlem} here.}

@@

@@

\begin{prop}
\label{prop:intsurj}
For $M \in M_n(\OK)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is $\OK[X] \cap K_{<n}[X]$.
\item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector.
\end{enumerate}
\end{prop}
\begin{proof}
The condition \eqref{pt:surjp} is equivalent to the surjectivity of $d\chi_M$ modulo $\pi$.  The equivalence with
\eqref{pt:cycp} follows the same argument as Proposition \ref{prop:surjectivity}, but over the residue field of $K$.
\end{proof}

Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$, and let
$\sigma_1(M), \dots, \sigma_n(M)$ denote the elementary divisors of $M$.

\begin{prop} \label{prop:mainlem}
Suppose that $M \in M_n(K)$ satisfies one of the conditions in Proposition \ref{prop:surjectivity}, and let 
\[
\alpha = \min\left(\prod_{i=1}^{n-1} \left\vert \sigma_i(M) \right\vert, 1\right).
\]
Then, for all $\rho \in (0, 1]$ and all
$r \in (0, \alpha^{-1} \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
satisfies:
\begin{equation}
\chi(M + H) = \chi(M) + d\chi_M(H).
\end{equation}
\end{prop}

\begin{proof}

Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
of $M$ is the lower convex hull of the Newton polygons of the entries of $\adj(X-M)$.
By \cite{caruso-roe-vaccon:15a}*{Prop. 3.4}, the endpoints of the precision polygon
occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
$B_W(1) \subset d\chi_M(B_V(\alpha^{-1}))$.

@@

\subsection{Stability under multiplication by $X$}

By definition, the codomain of $d \chi_M$ is $K_{< n}[X]$. 
However, when $M$ is given, $K_{< n}[X]$ is canonically isomorphic
to $K[X]/\chi_M(X)$ as a $K$-vector space. For our purpose, it will 
often be convenient to view $d \chi_M$ as an $K$-linear mapping
$M_n(K) \to K[X]/\chi_M(X)$.

\begin{prop}
Let $R$ be the subring of $K[X]$ consisting of polynomials $f$ for
which $f(M) \in M_n(\OK)$, and $\Lambda = d \chi_M \big(M_n(\OK)\big)$
as a submodule of $K[X]/\chi_M(X)$.  Then $\Lambda$ is stable under 
multiplication by $R$.
\end{prop}
\begin{proof}
Let $A = \adj(X-M)$ and $f \in R$.  By \eqref{eq:dchi}, $\Lambda$ is given by
the $\OK$-span of the entries of $A$.   Using the fact that the product of a matrix
with its adjugate is the determinant, $(X - M) \cdot A = \chi_M$ and thus
$f(X) \cdot A \equiv f(M) \cdot A \pmod{\chi_M(X)}$.  The span of the entries
of the left hand side is precisely $f(X) \cdot \Lambda$, while the span of the entries
of the right hand side is contained within $\Lambda$ since $f(M) \in M_n(\OK)$.
\end{proof}

@@

\subsection{Compact form of $d \chi_M$}

Let $\mathscr{C}$ be the companion matrix associated to $\chi_M$ as in \eqref{eq:companion}.
Assume one of the equivalent conditions of Proposition~\ref{prop:surjectivity}, giving a matrix 
$P \in \GL_n(K)$ such that $M = P \mathscr{C} P^{-1}$. Then Proposition~\ref{prop:surjectivity}
also holds for $M^T$, yielding an
invertible matrix $Q \in \GL_n(K)$ with $M^T = Q \mathscr{C} Q^{-1}$.

@@

\begin{rem}
With the previous notation, if $U \in GL_n(K),$
the quadruple for 
$UMU^{-1}$ is
$(\alpha, UP, (U^T)^{-1}Q, \chi_M),$
which can be computed in $O(n^\omega)$ operations in $K.$
This is faster than computing $U \adj(X-M) U^{-1},$
which is, at first sight, in $O(n^4)$\tristan{Shouldn't it be $\softO (n^4)$?} operations in $K.$
\end{rem}

>>>>

\newcommand{\famN}{\mathcal{N}}

@@

\item \done{In section 3, Algorithm 1, we have clarified why line 5 does not take quartic time}

@@

\item \done{The abstract has been rewritten. In particular, the kind of operations we are counted and the locution ``precision exactly $O(p^k)$'' have been clarified.}

@@

\item \done{The notion of lattice (in the tangent space) has been defined.}

@@

\done{
What we have done so far:
\begin{enumerate}
\item Section \emph{The contribution of this paper} of the introduction
has been rewritten and detailed. We split our contributions into two 
parts: the first part is about the (fast) computation of the adjugate
of $X{-}M$ while the second part concerns applications to precision.
We also added a sentence about numerical experiments.
\item Section 2.1 has been rewritten. We no longer talk about tangent
spaces. Jagged and flat precisions, together with their relationships with
lattice precision, are introduced and discussed. 
Proposition \ref{prop:mainlem} has been moved to Section 2.1 and stated
in terms of valuations.
\end{enumerate}
}

@@

Instead, we separate the computation of the precision of $\chi_M$ from the computation
of an approximation to $\chi_M$.  Given some precision on $M$, we use \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}
to find the best possible precision for $\chi_M$.  The analysis of this precision is the subject
of much of this paper.  With this precision known, the actual calculation of $\chi_M$
may proceed by lifting $M$ to a temporarily higher precision and then using a sufficiently
stable algorithm (see Remark \ref{rem:lift_for_optimal}).
\xavier{Maybe, we should say that it is not completely satisfying and (maybe) that we plan to come back to this question later.}

@@

We recall some of the definitions and results of 
\cite{caruso-roe-vaccon:14a} as a foundation for our discussion of the 
precision for the characteristic polynomial of a matrix. We will be 
concerned with two spaces in what follows: the space $M_n(K)$ of $n 
\times n$ matrices with entries in $K$ and the space $K_n[X]$ of monic 
degree $n$ polynomials over $K$. We define as well $K_{<n}[X]$ (resp. 
${\OK}_{<n}[X]$) as the space of polynomials over $K$ (resp. over $\OK$) 
of degree strictly less than $n$.

There are several choices for modeling precision on $p$-adic matrices. 
The first one consists in representing each entry of a matrix as a 
quantity of the form $a_{i,j} + O(\pi^{N_{i,j}})$. The integer $N_{i,j}$ 
is called the \emph{absolute precision} and may depend on $(i,j)$. The 
way of representing precision will be called \emph{jagged} in the 
sequel. When the $N_{i,j}$ are all equal (\emph{i.e.} when each entry 
is given at the same precision), we say that the precision datum is 
\emph{flat}.
A more flexible way of modeling precision consists in using lattices. We 
recall that a lattice in a $M_n(K)$ the image of $M_n(\OK)$ under a 
$\Qp$-linear transformation of $M_n(K)$. Since $\OK$ appears as the 
closed unit ball of $K$ (for the $p$-adic norm), cosets of the form
$M + H$ (where $M$ lies in $M_n(K)$ and $H$ is a lattice) are special
neighborhood of $M$; we will use them for modeling precision on $M$. 
Given a family of integers $\famN = (N_{i,j})_{i,j}$ the lattice:
$$H_{\famN} = 
\big\{ \, M = (a_{i,j})_{i,j} \in M_n(\Qp) \,\, | \,\, \forall i,\!j\!:
\pi^{N_{i,j}} \text{ divides } a_{i,j} \, \big\}$$
corresponds to the jagged precision introduced above.

Jagged and flat precisions are defining similarly for monic polynomials 
(taking care that no precision datum is needed for the leading 
coefficient which is supposed to be exactly~$1$). Simiarly, the notion 
of \emph{lattice precision} extends to polynomials. The underlying 
vector space is now $K_{<n}[X]$ and we define a lattice in it as the
image of ${\OK}_{<n}[X]$ under some linear mapping.

The benefit of working with lattices is that they behave well under 
differentiable maps; this is the content of 
\cite{caruso-roe-vaccon:14a}*{Lem. 3.4}. Write $\chi : M_n(K) \to 
K_n[X]$ for the characteristic polynomial map and $d \chi_M$ for its 
differential at $M$. Thanks to 
Recall from \citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} 
\cite{caruso-roe-vaccon:15a}*{\S 3.3}} that $d\chi_M$ is given by
\begin{equation} \label{eq:dchi}
\begin{array}{rcl}
d\chi_M: \quad M_n(K) & \to & K_{<n}[X] \smallskip \\
 dM & \mapsto & \tr(\adj(X{-}M) \cdot dM).
\end{array}
\end{equation}
where $\adj(M)$ denotes the adjugate\footnote{In 
\citelist{\cite{caruso-roe-vaccon:14a} \cite{caruso-roe-vaccon:15a}} the 
adjugate was referred to using the non-standard term \emph{comatrix}} of 
$M$ (when $M \in \GL_n(K)$, we have $\adj(M) = \det(M) M^{-1}$) and 
$d\chi_M$ the differential at $M$. In this setting, 
\cite{caruso-roe-vaccon:14a}*{Lem. 3.4} says that
\begin{equation}
\label{eq:mainlem}
\chi(M + H) = \chi(M) + d\chi_M(H)
\end{equation}
whenever $H$ is a sufficiently ``small'' and ``regular'' lattice.
More precisely, let $\sigma_1(M), \dots, \sigma_n(M)$ denote the 
elementary divisors of $M$.

\begin{prop} \label{prop:mainlem}
Let $M \in M_n(K)$ and assume that $d \chi_M$ is surjective. Let:
\[
\nu = \max\left(\sum_{i=1}^{n-1} \val(\sigma_i(M)), 0\right).
\]
Then, for all integers $N, M \geq 0$ with $M + \nu < 2N$
any lattice $H$ such that $M_n(\pi^N \OK) \subset H \subset
M_n(\pi^M \OK)$
satisfies Eq.~\eqref{eq:mainlem}.
\end{prop}

\begin{proof}
Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
of $M$ is the lower convex hull of the Newton polygons of the entries of $\adj(X-M)$.
By \cite{caruso-roe-vaccon:15a}*{Prop. 3.4}, the endpoints of the precision polygon
occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
${\OK}_{<n}[X] \subset d\chi_M(M_n(p^\nu \Zp))$.

\begin{comment}
Given a matrix $M \in M_n(K)$,
the most general kind of precision structure we may attach to $M$ is a
\emph{lattice} $H$ in the tangent space at $M$.  However, representing an
arbitrary lattice requires $n^2$ basis vectors, each with $n^2$ entries.  We therefore
frequently work with certain classes of lattices, either \emph{jagged} lattices
where we specify a precision for each matrix entry or \emph{flat} lattices where
every entry is known to a fixed precision $O(p^N)$.  Similarly, precision for
monic polynomials can be specified by giving a lattice in the tangent space
at $f(X) \in K_n[X]$, or restricted to jagged or flat precision in the interest
of simplicity.

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using 
\cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
For a matrix $M \in M_n(K)$, we identify the tangent space $V$
at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi_M$ with
the space $K_{<n}[X]$ of polynomials of degree less than $n$.
\end{comment}

@@

\noindent
{\bf About surjectivity.}
The surjectivity of $d \chi_M$ appeared as an assumption in 
Proposition~\ref{prop:mainlem}. The next Proposition make it
entirely explicit.

@@

\begin{prop}
\label{prop:intsurj}
For $M \in M_n(\OK)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is 
${\OK}_{<n}[X]$.
\item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector.
\end{enumerate}
\end{prop}
\begin{proof}
The condition \eqref{pt:surjp} is equivalent to the surjectivity of $d\chi_M$ modulo $\pi$.  The equivalence with
\eqref{pt:cycp} follows the same argument as Proposition \ref{prop:surjectivity}, but over the residue field of $K$.
\end{proof}

@@

\noindent
{\bf Stability under multiplication by $X$.}
By definition, the codomain of $d \chi_M$ is $K_{< n}[X]$. 
However, when $M$ is given, $K_{< n}[X]$ is canonically isomorphic
to $K[X]/\chi_M(X)$ as a $K$-vector space. For our purpose, it will 
often be convenient to view $d \chi_M$ as an $K$-linear mapping
$M_n(K) \to K[X]/\chi_M(X)$.

\begin{prop}
Let $R$ be the subring of $K[X]$ consisting of polynomials $f$ for
which $f(M) \in M_n(\OK)$, and $\Lambda = d \chi_M \big(M_n(\OK)\big)$
as a submodule of $K[X]/\chi_M(X)$.  Then $\Lambda$ is stable under 
multiplication by $R$.
\end{prop}
\begin{proof}
Let $A = \adj(X{-}M)$ and $f \in R$.  By \eqref{eq:dchi}, $\Lambda$ is given by
the $\OK$-span of the entries of $A$.   Using the fact that the product of a matrix
with its adjugate is the determinant, $(X{-}M) \cdot A = \chi_M$ and thus
$f(X) \cdot A \equiv f(M) \cdot A \pmod{\chi_M(X)}$.  The span of the entries
of the left hand side is precisely $f(X) \cdot \Lambda$, while the span of the entries
of the right hand side is contained within $\Lambda$ since $f(M) \in M_n(\OK)$.
\end{proof}

@@

\noindent
{\bf Compact form of $d \chi_M$.}
Let $\mathscr{C}$ be the companion matrix associated to $\chi_M$ as in \eqref{eq:companion}.
Assume one of the equivalent conditions of Proposition~\ref{prop:surjectivity}, giving a matrix 
$P \in \GL_n(K)$ such that $M = P \mathscr{C} P^{-1}$. Then Proposition~\ref{prop:surjectivity}
also holds for $M^T$, yielding an
invertible matrix $Q \in \GL_n(K)$ with $M^T = Q \mathscr{C} Q^{-1}$.

@@

\begin{rem}
With the previous notation, if $U \in GL_n(K),$
the quadruple for 
$UMU^{-1}$ is
$(\alpha, UP, (U^T)^{-1}Q, \chi_M),$
which can be computed in $O(n^\omega)$ operations in $K.$
This is faster than computing $U \adj(X-M) U^{-1},$
which is, at first sight, in $\softO(n^4)$
%\tristan{Shouldn't it be $\softO (n^4)$?}
operations in $K.$
\end{rem}
